Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-10 paper_io_sim.tex

    At any given time, a file can be partially cached, completely cached,
    or uncached at all. A cached data block can only reside in one of the two
    LRU lists. The first time they are accessed, blocks are
    added to the inactive list. On subsequent accesses, blocks of the
    inactive list are moved to the top of the active list. Cached blocks
    written to cache are marked as dirty until they are flushed to disk.

@@

            We first created a standalone prototype
            simulator to evaluate the accuracy and correctness of our
            model in a simple scenario before integrating it in the more complex
            WRENCH framework.
            The prototype uses the following basic storage model for
            both memory and disk:
            \begin{align*}
                & t_{r} = D / b_r \\
                & t_{w} = D / b_w\
            \end{align*}

@@

            We also implemented our model as part of WRENCH, enhancing its
            internal implementation and APIs with a page cache abstraction,
            and allowing users to activate the feature via a command-line
            argument. For the experiments in this paper we use WRENCH 1.6
            at commit
            \href{https://github.com/wrench-project/wrench/tree/67185374330d2c4bf274fce222c937e838df5b03}{6718537433},
            which uses SimGrid 3.25, available at
            \url{https://framagit.org/simgrid/simgrid}. Our implementation
            is now part of WRENCH's master branch and will be available to
            users with the next WRENCH release. WRENCH provides a full SimGrid-based simulation 
            environment that supports, among other features, concurrent accesses to storage devices, 
            applications distributed on multiple hosts, network transfers, 
            and multi-threading.

@@

        Our experiments compare real executions with our Python prototype,
        with the original WRENCH, and with our WRENCH-cache extension. We
        evaluate our page cache simulation in single-threaded and
        multi-threaded applications, accessing data on local and remote
        file systems. We use two applications: a synthetic one created to
        evaluate the simulation model, and a real one representative of
        neuroimaging data processing.

@@

        To parametrize our simulators, we benchmarked the
        memory, local disk, remote disk (NFS), and network bandwidths
        (Table~\ref{table:benchmark}). Since SimGrid, and thus WRENCH, currently only supports
        symmetrical bandwidths, we use the mean of the read and write
        bandwidth values in our experiments.
            \begin{table}[htbp]
            \centering
            \begin{tabularx}{\columnwidth}{|l
            |>{\centering\arraybackslash}X
            |>{\centering\arraybackslash}X
            |>{\centering\arraybackslash}X|}
            \hline
                Bandwidths  & Cluster (real) & Python simulator & WRENCH simulator\\
            \hline
                Memory read  & 6860    & 4812     & 4812\\
                Memory write & 2764    & 4812 & 4812\\
                Local disk read & 510 & 465 & 465\\
                Local disk write & 420 & 465     & 465\\
                Remote disk read & 515 & - & 445\\
                Remote disk write & 375 & - & 445\\
                Network & 3000 & - & 3000\\
            \hline
            \end{tabularx}
            \caption{Bandwidth benchmarks (MBps) and simulator configurations.
            The bandwidths used in the simulations are the average of the read and write bandwidths
            measured in the cluster.
            Remote disk and network accesses are not simulated in the Python simulator. \tristan{table could be prettified}}
            \label{table:benchmark}
            \end{table}

            As our focus is on I/O rather than compute, we measured
            application task CPU times on a cluster node
            (Table~\ref{table:cputime}), and used these durations in our
            simulations. For the Python simulator, we used CPU times
            directly in the simulation. For WRENCH and WRENCH-cache, we
            determined the corresponding number of flops on a 1~Gflops CPU
            and used these values in the simulation. The final simulated
            platform and application used in our experiments are avaiable at
            \url{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}.

@@

            The page cache model notably reduces WRENCH's simulation error
            for concurrent applications and local I/Os
            (Figure~\ref{fig:multi_local}). For reads, WRENCH-cache still
            slightly overestimates reality, due to the discrepancy between
            simulated and real read bandwidths mentioned before. For writes,
            WRENCH-cache slightly underestimates reality for the same reason. The plateau
            ending around 7 concurrent pipelines in the WRENCH-cache writes
            marks the limit beyond which the page cache is saturated with dirty data
            and needs flushing \tristan{why is there no plateau in real life?}.

@@

            Page cache simulation importantly reduces the simulation error
            on remote storage as well (Figure~\ref{fig:multi_nfs}). This
            manifests only for reads, as only writethrough cache was
            enabled in the NFS server. Both WRENCH and WRENCH-cache
            underestimate write times due to the discrepancy between
            simulated and real bandwidths mentioned previously. For reads,
            this discrepancy only impacts the results beyond 22
            applications since before this threshold, most reads result in cache
            hits \tristan{Dzung, pls check that.}.

>>>>

    At any given time, a file can be partially cached, completely cached,
    or not cached at all. A cached data block can only reside in one of two
    LRU lists. The first time they are accessed, blocks are
    added to the inactive list. On subsequent accesses, blocks of the
    inactive list are moved to the top of the active list. Cached blocks
    written to cache are marked as dirty until they are flushed to disk.

@@

            We first created a standalone prototype
            simulator to evaluate the accuracy and correctness of our
            model in a simple scenario before integrating it in the more complex
            \wrench framework.
            The prototype uses the following basic storage model for
            both memory and disk:
            \begin{align*}
                & t_{r} = D / b_r \\
                & t_{w} = D / b_w\
            \end{align*}

@@

            We also implemented our model as part of \wrench, enhancing its
            internal implementation and APIs with a page cache abstraction,
            and allowing users to activate the feature via a command-line
            argument. For the experiments in this paper we use \wrench 1.6
            at commit
            \href{https://github.com/wrench-project/wrench/tree/67185374330d2c4bf274fce222c937e838df5b03}{6718537433},
            which uses \simgrid 3.25, available at
            \url{https://framagit.org/simgrid/simgrid}. Our implementation
            is now part of \wrench's master branch and will be available to
            users with the next \wrench release. W\wrench provides a full \simgrid-based simulation 
            environment that supports, among other features, concurrent accesses to storage devices, 
            applications distributed on multiple hosts, network transfers, 
            and multi-threading.

@@

        Our experiments compare real executions with our Python prototype,
        with the original \wrench, and with our \wrench-cache extension. We
        evaluate our page cache simulation in single-threaded and
        multi-threaded applications, accessing data on local and remote
        file systems. We use two applications: a synthetic one created to
        evaluate the simulation model, and a real one representative of
        neuroimaging data processing.

@@

        To parametrize our simulators, we benchmarked the
        memory, local disk, remote disk (NFS), and network bandwidths
        (Table~\ref{table:benchmark}). Since \simgrid, and thus \wrench, currently only supports
        symmetrical bandwidths, we use the mean of the read and write
        bandwidth values in our experiments.
            \begin{table}[htbp]
            \centering
            \begin{tabularx}{\columnwidth}{|l
            |>{\centering\arraybackslash}X
            |>{\centering\arraybackslash}X
            |>{\centering\arraybackslash}X|}
            \hline
                Bandwidths  & Cluster (real) & Python simulator & \wrench simulator\\
            \hline
                Memory read  & 6860    & 4812     & 4812\\
                Memory write & 2764    & 4812 & 4812\\
                Local disk read & 510 & 465 & 465\\
                Local disk write & 420 & 465     & 465\\
                Remote disk read & 515 & - & 445\\
                Remote disk write & 375 & - & 445\\
                Network & 3000 & - & 3000\\
            \hline
            \end{tabularx}
            \caption{Bandwidth benchmarks (MBps) and simulator configurations.
            The bandwidths used in the simulations are the average of the read and write bandwidths
            measured in the cluster.
            Remote disk and network accesses are not simulated in the Python simulator. \tristan{table could be prettified}}
            \label{table:benchmark}
            \end{table}

            As our focus is on I/O rather than compute, we measured
            application task CPU times on a cluster node
            (Table~\ref{table:cputime}), and used these durations in our
            simulations. For the Python simulator, we used CPU times
            directly in the simulation. For \wrench and \wrench-cache, we
            determined the corresponding number of flops on a 1~Gflops CPU
            and used these values in the simulation. The final simulated
            platform and application used in our experiments are avaiable at
            \url{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}.

@@

            The page cache model notably reduces \wrench's simulation error
            for concurrent applications and local I/Os
            (Figure~\ref{fig:multi_local}). For reads, \wrench-cache still
            slightly overestimates reality, due to the discrepancy between
            simulated and real read bandwidths mentioned before. For writes,
            \wrench-cache slightly underestimates reality for the same reason. The plateau
            ending around 7 concurrent pipelines in the \wrench-cache writes
            marks the limit beyond which the page cache is saturated with dirty data
            and needs flushing \tristan{why is there no plateau in real life?}.

@@

            Page cache simulation importantly reduces the simulation error
            on remote storage as well (Figure~\ref{fig:multi_nfs}). This
            manifests only for reads, as only writethrough cache was
            enabled in the NFS server. Both \wrench and \wrench-cache
            underestimate write times due to the discrepancy between
            simulated and real bandwidths mentioned previously. For reads,
            this discrepancy only impacts the results beyond 22
            applications since before this threshold, most reads result in cache
            hits \tristan{Dzung, pls check that.}.
